---
title: 'Groundwater Flow Models in R Markdown: A Beginner''s Guide'
author: "Josh Clayton"
output:
  bookdown::gitbook: default
  bookdown::pdf_book: default
site: bookdown::bookdown_site
documentclass: book
---
#Objective

The objective of this guide is to introduce you to the modeling of groundwater flow, particularly the flow of near surface water. This guide will give you the basic tools you need to build a one-dimensional moisture profile using R Markdown. This guide is a simplified and reformatted version of the "Groundwater Flow" lab authored by Michael Dietze of Boston University's Earth and Environment Department.

#Background

##Intuition
At its core, this model describes a flux which can be mathematically expressed as $F_{k}=-k(\frac{dY}{dx})$ where *Y* is the variable of interest (i.e. concentration), *k* is the conductance, and *x* is distance. Specifically, this model is an application of [Darcy's Law](http://www.ldeo.columbia.edu/~martins/climate_water/lectures/darcy.html), which describes the flow of a liquid through a porous media.

It is important to note that conductivity declines as a function of depth; more compacted soils are less conductive. This model does not consider the gradient of conductivity for the sake of simplicity. Furthermore, this is a mere one-dimensional model that does not have a horizontal consideration. To best conceptualize the model you may want to envision a soil column that is 1 square meter on the surface of the soil.

##Model Variables
The following equations mathematically express the variables that compose this process-based model.

1. **Hydraulic conductivity** will be calculated as a function of soil moisture, $\theta_{i}$, for each depth, i.
$$k_{i,k} = k_{sat} \frac{ \theta_{i,t}}{\theta_{sat}}^{2b+3} $$
2. **Water potential** will be calculated as the sum of *gravitational potential* and the *matric potential*, illustrated below.

- gravitational potential: $\psi_{g,i}=h_{i}$
- matric potential: $\psi_{m,i,t}=\psi_{sat}\frac{\theta_{i,t}}{\theta_{sat}}^{-b}$
- **water potential**: $$\psi_{i,t}=\psi_{m,i,t}+\psi_{g,i}$$ 

3. **Flux Rate** will be calculated from the difference in *water potential*.
$$F_{i,t} = -k_{i,t}\frac{d\psi_{i,t}}{dz} $$
$$Vol_{i,t}=min(sink.max,source.max,flux_{i,t}(dt)) $$
4. **Soil moisture** in this discrete model will be calculated by updating the soil moisture for the next time step based on the current moisture fluxes both *in* and *out* of each previous layer.
$$\theta_{i,t+1}=\theta_{i,t}+Vol_{i-1,t}-Vol_{i,t}$$

#Data
Download and save the **texture.csv** data file provided in the repository if you have not already. This file provides a list of the physical properties of soils of different textures. Ensure that both the .Rmd file that you are building your model in and the **texture.csv** data file are both saved within the save folder that you have set as your [working directory](http://rfunction.com/archives/1001).

```{r}
dat = read.csv("texture.csv",header=TRUE)
```

It may prove useful to survey the data. In this example we will simulate groundwater flow through **loamy sand**.

- **slpots**: soil moisture potential at saturation (m)
- **slmsts**: soil moisture at saturation ($\frac{m^3}{m^3}$)
- **slcons0**: surface value of hydraulic conductivity at saturation ($\frac{m}{s}$)
- **slbs**: B exponent ($\frac{n}{d}$) *which you saw earlier in the equation for hydraulic conductivity*
- **soilcp**: dry soil capacity (-3.1MPa) ($\frac{m^3}{m^3}$)
```{r}
knitr::kable(dat[,c("name","slpots","slmsts","slcons0","slbs","soilcp")])
```

#Model

##Initial Conditions
This model will be simulated on a soil profile **2 meters deep**  with **8 layers** that are each **0.25 meters** thick. Other than defining a depth that is shallow (only 2 meters below the surface), these conditions were arbitrarily assigned. Feel free to assign your own, especially if it renders your model more informative.
```{r}
z=2 #depth (m)
n=8 #number of soil layers
dz=0.25 #layer thickness (m)
```

Additionally, this model will be simulated over a period of **36 hours** with individual, discrete time steps of **10 seconds**.
```{r}
p=36 #period
dt=10 #seconds
nt=p*3600/dt #time steps
```

Below we define soil as layers 2 through 9 because the top, or first, layer is above ground (and thus not soil). Defining which layers constitute soil will serve as a useful indexing tool later on.
```{r}
soil=2:9
```


##Setup
Below, matrices are built to manage the variables described in the background section. The model is initialized with a $\theta_{i}$ of 0.3 at each level.
```{r}
theta=matrix(0,nrow=(n+1),ncol=nt+1) #moisture matrix
theta[2:9,1]=.3 #initial moisture conditions
k=matrix(0,nrow=n,ncol=nt) ##hydrualic conductivity (k) matrix
ksat=dat[2,"slcons0"] #hydraulic conductivity at saturation
thetasat=dat[2,"slmsts"] #saturation 
b=dat[2,"slbs"] #B exponent value
h=c(0,-.25,-.5,-.75,-1,-1.25,-1.5,-1.75,-2) #depth profile
gpot=h #has length 9
psisat=dat[2,"slpots"] #moisture at saturation
psi=matrix(0,nrow=n+1,ncol=nt) #potential matrix
flux=matrix(0,nrow=n,ncol=nt) ##flux
dry=dat[2,"soilcp"] #dry soil moisture
vol=matrix(0,nrow=n,ncol=nt) #volume flux
```

##The Model
This model loops over time and depth for each time step.
```{r}
for(t in 1:nt){
  for(i in 1:n){ #hydraulic conductivity
    k[i,t]=ksat*(theta[i+1,t]/thetasat)^((2*b)+3)
  }
  psi[1,t]=theta[1,t] #water potential
  for(i in 2:(n+1)){ 
    psi[i,t]=h[i]+psisat*(theta[i,t]/thetasat)^(-b)
  }  
  for(i in 1:n){ #flux
    flux[i,t]=-k[i,t]*((psi[i+1,t]-psi[i,t])/dz)
  }
  for(i in 1:n){ #volume flux
    if(i==1){
      source.max=theta[1,t]
    }else{
      source.max=theta[i,t]-dry
    }
    sink.max=thetasat-theta[i+1,t]
    vol[i,t]=max(0,min(sink.max,source.max,flux[i,t]*dt))
  }
  theta[1,t+1]=theta[1,t]-vol[1,t]
  for(i in 2:n){
    theta[i,t+1]=theta[i,t]+vol[i-1,t]-vol[i,t]
  }
  theta[n+1,t+1]=theta[n+1,t]+vol[n,t]
}
```


#Graphical Representation
##Soil Moisture over Time
```{r}
plot(theta[2,],
     ylim=c(.1,.5),
     xlim=c(1,200000),
     xlab="Time (sec)",
     ylab="Soil Moisture (m^3/m^3)",
     type='l',
     col=2,
     lwd=2,
     main="Soil Moisture vs. Time",
     log='x')
for(i in 3:(n+1)){
   lines(theta[i,],col=i,lwd=2)
}
abline(h=thetasat,lty=2)
legend("topright",legend=paste(h[2:9],"m"),col=2:9,lwd=2)
```

##Steady State
Below we calculate the time at which the model in this simulation reached [steady state](http://groups.molbiosci.northwestern.edu/holmgren/Glossary/Definitions/Def-S/steady_state.html).
```{r}
steadystate=rep(0,9)
for(i in 1:9){
  steadystate[i]=which(round(theta[i,],digits=3)==round(theta[i,nt],digits=3))[1]
}
max(steadystate)
```
Below is a graph of the soil moisture at each level at *steady state*.
```{r}
soil = 2:(n+1)
plot(-h[soil],theta[soil,nt+1],
     ylab="Soil Moisture (m^3/m^3)",
     xlab="Depth (m)",
     main="Soil Moisture vs. Depth at Steady State"
     )
```


#Additional Applications
##Surface Flood Simulation
One reason an above ground layer was defined was to allow for simulations of surface flood.
```{r}
theta[1,1]=.3
for(t in 1:nt){
  for(i in 1:n){
    k[i,t]=ksat*(theta[i+1,t]/thetasat)^((2*b)+3)
  }
  psi[1,t]=theta[1,t]
  for(i in 2:(n+1)){ 
    psi[i,t]=h[i]+psisat*(theta[i,t]/thetasat)^(-b)
  }  
  for(i in 1:n){
    flux[i,t]=-k[i,t]*((psi[i+1,t]-psi[i,t])/dz)
  }
  for(i in 1:n){
    if(i==1){
      source.max=theta[1,t]
    }else{
      source.max=theta[i,t]-dry
    }
    sink.max=thetasat-theta[i+1,t]
    vol[i,t]=max(0,min(sink.max,source.max,flux[i,t]*dt))
  }
  theta[1,t+1]=theta[1,t]-vol[1,t]
  for(i in 2:n){
    theta[i,t+1]=theta[i,t]+vol[i-1,t]-vol[i,t]
  }
  theta[n+1,t+1]=theta[n+1,t]+vol[n,t]
}
```
```{r}
plot(theta[2,],ylim=c(.15,thetasat),xlim=c(0,16500),xlab="Time (sec)",ylab="Soil Moisture (m^3/m^3)",type='l',col=2,main="Soil Moisture vs. Time",lwd=3)
for(i in 3:(n+1)){
   lines(theta[i,],col=i,lwd=3)
}
legend("topright",legend=paste(h[2:9],"m"),col=2:9,lwd=2)
```

